Are populations like a circuit? Comparing isolation by resistance to a new coalescent-based method.

### 2019-03-02 review-responses.tex

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript, 
``Are populations like a circuit? 
Comparing isolation by resistance to a new coalescent-based method.''
(formerly titled ``Are populations like a circuit? The relationship between isolation by distance and isolation by resistance'').

@@

\reply{
    In the current implementation, 
    equation (7) is solved using the solve(A,b) in R. 
    The actual method varies depending on the specific linear algebra library used, 
    but it is typically Gaussian elimination or LU decomposition, 
    both of which are $\mathcal{O}(N^3)$ 
    (unfortunately the decomposition is different each step;
    if it were the same it would be $\mathcal{O}(N^2)$).
    Since $n$ demes gives you order $n^2$ equations and unknowns, 
    the time complexity is 
}

\begin{point}{}
    The title of the manuscript is a bit misleading. One would not guess that
    the introduction of a method to infer biased and asymmetric gene flow is the
    main point.  ``Isolation by distance'' does not appear often in the manuscript,
    neither its classic literature (Wright, Maryuama, Malecot, Barton etc.) nor
    direct investigation of genetic differentiation with geographic distance. But
    these things are also not the scope of this paper.  
\end{point}

% Citation:
% The Effects of Genetic and Geographic Structure on Neutral Variation
% Brian Charlesworth, Deborah Charlesworth and Nicholas H. Barton
% Annual Review of Ecology, Evolution, and Systematics
% Vol. 34 (2003), pp. 99-125

\reply{
    We do think a big point of the manuscript is discussion of the drawbacks of resistance-based methods,
    but we agree with the point about isolation by distance.
    We've retitled it: 
    ``Are populations like a circuit? 
    Comparing isolation by resistance to a new coalescent-based method.''
}

@@

\begin{point}{}
    You could write an outlook stating that your method could be used to fit
    parameterized models with few parameters. For instance, in case the putative
    location of a barrier or an area of biased gene flow is known, one could try to
    infer it without fitting ALL movement parameters, but a simple parametric
    model, which would implicitly take care of the regularization.  
\end{point}

\reply{
    This is a good thing to emphasize. We've added a bit to this effect. \revref
}

@@

\begin{point}{}
    You have your own implementation of resistance distance methods, by first
    extracting hitting times. However, you allow non-symmetric rates, something
    no existing resistance method would do. Given your observation that there
    is a lot degeneracy for resistance times (adding skew stuff to hitting
    times), this introduces uncertainty that likely cannot be broken such as
    with coalescence methods. This could drive the high uncertainty in Fig 5
    upper right. Given your good results for Fig. 4 for the symmetric case,
    this is a bit surprising. Maybe run one scheme where you FIX to a single
    symmetric movement rate between each pair of demes. This setting would be a
    fairer comparison (and be much closer to what EEMS etc. actually do -- they
    do not even try to infer asymmetric rates).
\end{point}

\reply{
    This is something we were considering doing. 
    We decided to allow symmetric migration in the resistance distance case 
    in order to make the two methods as similar as possible, 
    although it could arguably be more useful to compare them
    when they are implemented how they would be in practice.
}

\begin{point}{}
    Maybe describe the model you use for you calculation and inference, a
    continuous backward in time Markov model with coalescence rates in demes
    before the actual recursions, and not bit by bit in there.
\end{point}

\reply{
    We've added a bit more to our big-picture overview that we hope achieves
    the goal of setting the stage appropriately. \revref
}

\begin{point}{\revref}
    Do you actually use this tensor product for the movement
    rates anywhere? And seems Eq.\ 7 (which has no tensor product) is all you
    need.
\end{point}

@@

\reply{
    Thank you for catching this.
    You are correct that in addition to $ZC$ being skew-symmetric,
    the rows of $Z$ must also sum to zero.
    The second part is also a good point, 
    but probably beyond the scope of this paper.
}

\begin{point}{Eq. 4:}
    $H$ may not be always invertible ($H$ symmetric and positive off-diagonal
    and 0 on the diagonal is NOT enough). So there might be some pathological
    cases where multiple G link to one $H$, maybe even in some simple toy
    examples.
\end{point}

\reply{
    Good point; we had not justified this.
    Under the standing assumption that the chain is irreducible, 
    $H$ is always invertible.
    We now prove this in the Appendix \llname{rr:H_invertible},
    and have made reference to this fact. \revref
}

\begin{point}{}
    Dimensionality arguments (Line 129 onward) can be a bit misleading when you
    already know the equations have a non-trivial kernel. Setting all
    non-neighboring elements of $G$ to 0 might not get rid of parts of the
    kernel. So maybe flag it even more as ``be cautious'' as you already have.
\end{point}

\reply{
    Good point; we've changed the wording here. \revref
}

@@

\begin{point}{}
    You could just call the thing by its name: ``Stepping stone'' model. This
    term has such a rich history in population genetics!
\end{point}

\reply{
    In our experience, the term ``stepping stone model'' is used in a variety of ways in the literature
    (e.g., also for all-migrate-to-all island models),
    so we prefer to describe explicitly the grid that we use.
}

@@

\begin{point}{}
    In the main text, you never mention that the mean of the exponential
    movement rates for simulations is one, only in a figure. One can also kind
    of see it from the model figures, but it would be nice to have it
    explicitly stated in the main text.
\end{point}

\reply{
    Good point; we've added this. \revref
}

\begin{point}{}
    In the figures, probably keep model figure and inference results closely
    together. This helps the reader, and makes the figures stand more for
    themselves.
\end{point}

\reply{
    This is a good idea,
    but we haven't been able to see how to implement in a way that would improve things.
    Maybe we don't see your idea,
    but the additional graphs that describe population structure in the Appendix
    won't be legible if added to existing figures,
    and we think we already have more than enough figures.
}

@@

\begin{point}{}
    Useful result to know: Maryuama’s invariance principle (Nagylaki 1998):
    That the correctly weighted within coalescence time within demes only
    depends on the total population size  (given conservative migration) -- so
    it cannot fluctuate too wildly across demes and movement rates! This fact
    motivates why not fitting too wildly differing within-deme coalescence
    times/diversities.
    % Nagylaki, T., 1998 The expected number of heterozygous sites in a subdivided
    % population. Genetics 149: 1599
\end{point}

\reply{
    Good reference! We'd referred to the simpler version of \citet{strobeck1987average},
    and added this reference. \revref
}

\begin{point}{\revref}
    400,000 generations will look a bit lunatic for someone not in
    popgen who does not know about rescaling. You could give re-scaled, more
    realistic values!
\end{point}

\reply{
    We didn't really intend that number to sound so outlandish,
    but we think it's more ``quite long'' than ``absurdly long''?
    We're not sure that mentioning rescaling will help,
    but have added a reassuring note. \revref
    (Note that we \emph{did} actually simulate for that many generations;
    we didn't use rescaling to run a shorter or smaller simulation.)
}

@@

\begin{point}{Paragraph beginning \revref:}
    Maybe move this section in front of the inference part, as it is useful to
    know the most ``fundamental'' limit of resistance distances.
\end{point}

\reply{
    This is a good point, but we'd rather keep the different results having to do with the same model
    together in the same place.
}

@@

\begin{point}{Fig. 7:}
    With EEMS one can see very suspiciously the 4x4 sampling structure - there
    are red dots in almost all 3x3 centers of the sampling squares. Since EEMS
    is quite fast, could you increase the resolution of sampling?
\end{point}

\reply{
    More samples could be added for the EEMS runs, 
    but doing so would have the drawback 
    that the two methods would no longer be using the same data for inference. 
    We've been told that similar patterns show up when EEMS tries to approximate
    anisotropic migration,
    so it may not be the sampling pattern.
    We've added a note about this. \revref
    (Also, note that these samples are more spread out that people often have in practice:
    they do occur in the middle three-quarters of each grid square,
    but they aren't all located at the center.)
}

@@

\begin{point}{}
    ``Bayesian inference of movement rates'': In their model the authors assume
    that many entries of the migration rate matrix are zero, and place an
    exponential prior on the rest. Is it safe to assume that we will always
    want to constrain rates between discontiguous cells to zero? Also, given
    the valid concerns that they raise elsewhere about identifiability and
    power, I wonder if the authors considered placing a sparse prior on the
    entire G matrix, enabling the model to adapt to the amount of signal while
    shrinking other migration rates to zero. This might have the effect of
    making the resulting non-zero estimates more stable. (I realize that this
    raises potential problems for inference if the underlying graph becomes
    disconnected.)
\end{point}

\reply{
    Whether or not it is reasonable to assume there is no long distance migration
    depends on the population under study. 
    There are cases where it might be a good idea to allow it.
    Putting a sparse prior on all of $G$ is a good idea for future improvements, 
    although it would probably be appropriate for the prior probability of a value
    to be non-zero to be much higher for adjacent demes as compared to non-adjacent ones, 
    which would also likely prevent the graph from becoming disconnected very often.
    Since disconnecting the graph would result in some infinite coalescence times, 
    these steps would be rejected anyway.
}

\begin{point}{}
    The discussion of ill-conditioning is welcome. From experience, it often
    feels like coalescent-based methods have little power to accurately
    estimate migration unless the rates are large. It might be interesting to
    shift emphasis from estimation to the slightly easier problem of testing:
    how good are these methods at distinguishing between a given null and an
    alternative hypothesis (say, barrier vs.\ no barrier) for the rate matrix $G$?
    I think the gain of using a coalescent model might be more apparent in this case.
\end{point}

\reply{
    This is a good point; we've added it to the discussion
    (and credited you with it). \revref
}

### 2019-03-02 isolation_by_coalescence.tex
responses during meeting

\title{Are populations like a circuit? The relationship between isolation by distance and isolation by resistance}
\author{Erik Lundgren\footnote{Computational Biology, University of Southern California}, and
        Peter L. Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
\maketitle

@@

The \emph{commute time} of a Markov chain between locations $i$ and $j$
is the expected time for a particle that moves according to the rules of the chain, 
started at $i$, to first reach $j$ and then return.
We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
between $i$ and $j$.
The \emph{coalescence time} is the expected time for two particles started at $i$ and $j$
to \emph{coalesce},
which happens at a given rate whenever they are in the same location.
We will denote this by $C_{ij}$.
Why should these be related to genetic distance? 
First, we explain this connection,
and then discuss their computation and application.

@@

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous-fill-2014}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
(We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.)
Therefore, $H$ uniquely determines $G$.

@@

Since coalescence times are symmetric, 
equation \eqref{eqn:C_matrix} provides $n (n+1)/2$ informative equations.
This is larger than $4n+n$, the number of parameters,
for a grid with at least nine nodes (i.e., $3 \times 3$),
so the system of equations will usually have a unique solution
for grids larger than this, except in special cases.
The same calculation holds for commute times 
except that there are $n$ local diversity parameters instead of $n$ coalescence parameters.

@@

@@

We first tested the method using data generated under the model.
To test the methods across a wide range of situations,
we generated fifty sets of random movement rates on $3 \times 2$ rectangular grids:
twenty-five symmetric (i.e., with $G_{ij} = G_{ji}$) and twenty-five asymmetric.
We generated movement parameters
by rounding independent exponential random variables up to the nearest one tenth,
and fixed coalescence rates at 1 for all locations.
We then analytically computed mean coalescence times,
and added independent noise to each pairwise coalescence time.
We assume that we can estimate the amount of noise reasonably well,  \revpoint{3}{4}
so used the true standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
To assess the impact on inference of the number of parameters,
we performed inference both with and without the assumption 
that coalescence rate is the same everywhere.
For $3 \times 2$ graphs, there are 21 equations
and 14 movement parameters,
so that with a global coalescence rate there are 15 unknowns.
However, with individual coalescence rates there are 20 unknowns,
which is close to nonidentifiable.
To quantify accuracy across these random graphs in different situations,
we computed, for each model fit on a particular random graph,
the absolute difference between the posterior median of each movement parameter
and the true value.
We then averaged these across movement parameters to produce a measure of accuracy for that model fit,
reported below as ``mean absolute error''.

@@

\paragraph{Other barriers: population sinks and recent expansions}
We simulated two other situations on a continuous landscape that produced an bias in effective migration.
First, we simulated a \textbf{valley} in a landscape with a $5 \times 3$ aspect ratio as above, 
in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
The valley is a population ``sink'',
although it would not be detectable with mortality rates,
and shows only a slightly lower population density (Supplemental Figure \ref{sfig:valley_results}).
Since individuals in the valley are more likely to have parents outside of the valley,
this produces gene flow ``uphill'', away from the ridge.
Next, we simulated a recent \textbf{expansion},
in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations, \revpoint{1}{18}
until 500 generations before the present day
(a model of secondary contact).
The resulting expansion produces gene flow back away from the center,
similar to the ``valley''.

Both coalescence time inference and the resistance landscape inferred by EEMS
show a barrier to gene flow in the middle of the landscape (Figure \ref{fig:more_barriers}).
This is arguably correct, 
although there is in fact no current barrier to migration in either simulation.
Coalescence time inference also correctly infers 
the general bias in recent gene flow away from the center,
although as in other continuous-space simulations, 
there was substantial noise in the inferred patterns.
However, the resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance.
These patterns are commonly seen when attempting to use EEMS, 
whose underlying model of gene flow is symmetric,
to describe anisotropic gene flow.
Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
and our implementation of resistance distance inference,
which allows asymmetric gene flow, shows similar patterns
(Supplemental Figures \ref{sfig:valley_results} and \ref{sfig:expansion_results}).

@@

Inference of movement rates, in either a coalescence time or resistance distance framework,
is an ill-conditioned problem,
implying the need for regularization to obtain reliable inference.
This fact could explain the results of \citet{graves2013current},
who found a large area of resistance parameters
that produced equivalently good fits to genetic distance
(i.e., nonidentifiability, analogous to a flat likelihood surface).

@@

Can we solve equation \eqref{eqn:R_eq_C} more generally?
For that equation to hold, we need $Gq \bone^T + \bone (Gq)^T$ to be diagonal,
i.e., that $(Gq)_i + (Gq)_j = 0$ for all $i \neq j$.
For any $k \neq \ell$, there exists a vector $u^{(k\ell)}$ such that $(Gu^{(k\ell)})_i = \delta_{ik} - \delta_{i\ell}$;
the only possible $q$ for which $Gq \bone^T + \bone (Gq)^T$ is diagonal
are of the form $q = \alpha \bone + \beta u^{(k,\ell)}$ for some $k \neq \ell$ and some constants $\alpha$ and $\beta$.
We would then need
\begin{align}
    1/\pi_i + \delta_{ik} - \delta_{i\ell} = q_i \gamma_i \qquad \text{for all} \quad i.
\end{align}
This implies that for any Markov chain with symmetric hitting times,
we can find diversity values ($q$) and coalescence rates ($\gamma$) that make $\comdist = C$
(and in fact there are many ways to do this).
However, there is not a general solution for $q$ if coalescence rates are also given (as is the case in practice).
This is related to the fact proved by \citet{strobeck1987average} 
that $\diag(C)$ is constant and does not depend on movement rates
for any isotropic conservative migration model.

>>>>

\title{Are populations like a circuit? 
Comparing isolation by resistance to a new coalescent-based method.}
\author{Erik Lundgren\footnote{Computational Biology, University of Southern California}, and
        Peter L.\ Ralph\footnote{Institute for Ecology and Evolution, University of Oregon}}
\maketitle

@@

The \emph{commute time} of a Markov chain between locations $i$ and $j$
is the expected time for a particle that moves according to the rules of the chain, 
started at $i$, to first reach $j$ and then return.
We will denote this by $R_{ij}$, to emphasize that it is also the effective resistance
between $i$ and $j$.
The \emph{coalescence time} is the expected time for two particles started at $i$ and $j$
to \emph{coalesce},
which happens at a given rate whenever they are in the same location.
We will denote this by $C_{ij}$.
Why should these be related to genetic distance? 
First, we explain this connection,
and then discuss their computation,
and finally develop a method that infers movement rates of the Markov chain
(which are reverse-time migration probabilities)
from observations of coalescence time. \revpoint{1}{9}

@@

Suppose instead we are given the hitting times, $H$, and wish to find the movement rates, $G$.
This is not the situation we are in -- we have either commute or coalescence times --
but it is related.
The Random Target Lemma \citep{aldous-fill-2014}
tells us that the stationary distribution of the Markov chain, denoted $\pi$,
can be recovered from the hitting times by solving $\pi = H^{-1} \bone / \bone^T H^{-1} \bone$.
As shown in Appendix \ref{apx::hitting_calcs},
equation \eqref{eqn:hitting_sum} can be rewritten in matrix form as
\begin{align}
    G H = \diag(1/\pi) - \bone \bone^T ,
\end{align}
which implies that $G$ can be computed directly from $H$ as
\begin{align} \label{eqn:G_from_H}
    G = \left( \diag(1/\pi) - \bone \bone^T \right) H^{-1} .
\end{align}
(We prove that $H$ is invertible in Appendix~\ref{apx::hitting_calcs}.) \revpoint{1}{12}
Therefore, $H$ uniquely determines $G$.

@@

Since coalescence times are symmetric, 
equation \eqref{eqn:C_matrix} provides $n (n+1)/2$ informative equations.
This is larger than $4n+n$, the number of parameters,
for a grid with at least nine nodes (i.e., $3 \times 3$),
so we would expect the system of equations to usually have a unique solution
for grids larger than this,
although some special cases will still have free parameters. \revpoint{1}{13}
The same calculation holds for commute times 
except that there are $n$ local diversity parameters instead of $n$ coalescence parameters.

@@

One strategy for circumventing ill-conditioning is to impose additional constraints --
for instance, constraining all population sizes to be the same (something we do below),
or allowing only two distinct migration rates: one rate across a barrier, 
and another across ``non-barriers''.
Such constraints amount to a form of regularization,
since they look for solutions to the original problem with certain desireable properties
(e.g., most of the values are the same). \revpoint{1}{4}

@@

We first tested the method using data generated under the model.
To test the methods across a wide range of situations,
we generated fifty sets of random movement rates on $3 \times 2$ rectangular grids:
twenty-five symmetric (i.e., with $G_{ij} = G_{ji}$) and twenty-five asymmetric.
We generated movement parameters
by rounding independent exponential random variables with mean 1 up to the nearest one tenth, \revpoint{1}{15}
and fixed coalescence rates at 1 for all locations.
We then analytically computed mean coalescence times,
and added independent noise to each pairwise coalescence time.
We assume that we can estimate the amount of noise reasonably well,  \revpoint{3}{4}
so used the true standard deviation of the noise for $\sigma_\epsilon$ in the likelihood function.
To assess the impact on inference of the number of parameters,
we performed inference both with and without the assumption 
that coalescence rate is the same everywhere.
For $3 \times 2$ graphs, there are 21 equations
and 14 movement parameters,
so that with a global coalescence rate there are 15 unknowns.
However, with individual coalescence rates there are 20 unknowns,
which is close to nonidentifiable.
To quantify accuracy across these random graphs in different situations,
we computed, for each model fit on a particular random graph,
the absolute difference between the posterior median of each movement parameter
and the true value.
We then averaged these across movement parameters to produce a measure of accuracy for that model fit,
reported below as ``mean absolute error''.

@@

\paragraph{Other barriers: population sinks and recent expansions}
We simulated two other situations on a continuous landscape that produced an bias in effective migration.
First, we simulated a \textbf{valley} in a landscape with a $5 \times 3$ aspect ratio as above, 
in which fecundity of individuals in the middle 1/4 of the landscape was reduced by 50\%.
The valley is a population ``sink'',
although it would not be detectable with mortality rates,
and shows only a slightly lower population density (Supplemental Figure \ref{sfig:valley_results}).
Since individuals in the valley are more likely to have parents outside of the valley,
this produces gene flow ``uphill'', away from the ridge.
Next, we simulated a recent \textbf{expansion},
in which the middle 4/5 of the landscape was uninhabitable for 400,000 generations, \revpoint{1}{18}
until 500 generations before the present day.
(This is a model of secondary contact with a long period of isolation,
but not outside the realm of possibility.)
The resulting expansion produces gene flow back away from the center,
similar to the ``valley''.

Both coalescence time inference and the resistance landscape inferred by EEMS
show a barrier to gene flow in the middle of the landscape (Figure \ref{fig:more_barriers}).
This is arguably correct, 
although there is in fact no current barrier to migration in either simulation.
Coalescence time inference also correctly infers 
the general bias in recent gene flow away from the center,
although as in other continuous-space simulations, 
there was substantial noise in the inferred patterns.
However, the resistance landscapes as inferred by EEMS show patchy, vertical bands of high resistance.
These patterns are commonly seen when attempting to use EEMS, 
whose underlying model of gene flow is symmetric,
to describe anisotropic gene flow.
The grid-like pattern visible in this and other EEMS results
may be due to the sampling scheme
(recall that samples are drawn from the middle three-quarters of each grid square). \revpoint{1}{20}
Our analysis above shows that this is an unavoidable consequence of the resistance approximation,
and our implementation of resistance distance inference,
which allows asymmetric gene flow, shows similar patterns
(Supplemental Figures \ref{sfig:valley_results} and \ref{sfig:expansion_results}).

@@

Inference of movement rates, in either a coalescence time or resistance distance framework,
is an ill-conditioned problem,
implying the need for regularization to obtain reliable inference.
This fact could explain the results of \citet{graves2013current},
who found a large area of resistance parameters
that produced equivalently good fits to genetic distance
(i.e., nonidentifiability, analogous to a flat likelihood surface).
An anonymous reviewer suggested
that although inference of landscapes of movement rates may carry a degree of unavoidable uncertainty,
the problem of \emph{model comparison} 
(e.g., between models with and without migration across a certain barrier)
may still produce more certain answers. \revpoint{2}{2}

@@

Can we solve equation \eqref{eqn:R_eq_C} more generally?
For that equation to hold, we need $Gq \bone^T + \bone (Gq)^T$ to be diagonal,
i.e., that $(Gq)_i + (Gq)_j = 0$ for all $i \neq j$.
For any $k \neq \ell$, there exists a vector $u^{(k\ell)}$ such that $(Gu^{(k\ell)})_i = \delta_{ik} - \delta_{i\ell}$;
the only possible $q$ for which $Gq \bone^T + \bone (Gq)^T$ is diagonal
are of the form $q = \alpha \bone + \beta u^{(k,\ell)}$ for some $k \neq \ell$ and some constants $\alpha$ and $\beta$.
We would then need
\begin{align}
    1/\pi_i + \delta_{ik} - \delta_{i\ell} = q_i \gamma_i \qquad \text{for all} \quad i.
\end{align}
This implies that for any Markov chain with symmetric hitting times,
we can find diversity values ($q$) and coalescence rates ($\gamma$) that make $\comdist = C$
(and in fact there are many ways to do this).
However, there is not a general solution for $q$ if coalescence rates are also given (as is the case in practice).
This is related to the fact proved by \citet{strobeck1987average} 
(and generalized by \citet{nagylaki1998expected}) \revpoint{1}{17}
that $\diag(C)$ is constant and does not depend on movement rates
for any isotropic conservative migration model.
